<!-- rmarkdown v1 -->

---
title: "PA1template.MD"
author: "Rod Maclean"
date: "January 29, 2016"

---
## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement -- a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.
## Data

This data set was sourced from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
 at 2016-01-29 09:31:05 EST

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Analysis Processing


The following libraries may be required:

```{r}
##load libraries
library ("data.table")
library ("tidyr")
library("dplyr")
library("lubridate")

```

First the data set is loaded into R

```{r}
##read data files
data_set <- fread("./activity.csv", header = TRUE,sep="," , stringsAsFactors =FALSE)
```
We then clean up the data 

```{r}
##Convert date to POSIXct
data_set$date<-ymd(data_set$date)
##Convert interval to factor
data_set$interval <-as.factor(data_set$interval)
```

## What is mean total number of steps taken per day?

A summary of the number of steps per day is created, ignoring missing data
```{r}
#Summarize to steps per day
steps_per_day <-summarize(group_by(data_set,date),total_steps =sum(steps, na.rm = TRUE))
```

A histogram of the number of steps per day is produced. 

Notice the large number of days with no activity. This is an artifact of the missing data
```{r}
#Histogram of number of steps per day
with(steps_per_day , hist(total_steps, breaks=10))
```

The mean number of steps per day is calculated as below:
```{r}
#Mean of Number of total steps per day
summarize(steps_per_day,mean_steps =mean(total_steps, na.rm = TRUE))
```
The median number of steps per day is calculated as below:
```{r}
#Median Number of total steps per day
summarize(steps_per_day,median_steps =median(total_steps, na.rm = TRUE))
```
## What is the average daily activity pattern?

### This is a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
#Daily Activity Plot
steps_per_interval <- summarize(group_by(data_set,interval),mean_steps =mean(steps, na.rm = TRUE))
with(steps_per_interval , plot(interval, mean_steps, type = "n", xlab="Time Interval Number", ylab="Average Number of Steps"))
lines(steps_per_interval$interval, steps_per_interval$mean_steps, col="blue",type ="S")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
#Interval with maximum steps
steps_per_interval[steps_per_interval$mean_steps==max(steps_per_interval$mean_steps)]
```

```{r}
#Number of intervals with no step data
sum(is.na(data_set$steps))
```
## Imputing missing values

 The chosen strategy for replacing missing values was to replace the missing value with the rounded mean value for that interval.  
```{r}
#Calculate rounded mean steps per interval
steps_per_interval <- summarize(group_by(data_set,interval),mean_steps =round(mean(steps, na.rm = TRUE)))

#Add data to data set
filled_set <- copy(data_set)
filled_set$filled_steps  <-ifelse(is.na(data_set$steps)==TRUE,steps_per_interval[match(data_set$interval,steps_per_interval$interval),]$mean_steps,data_set$steps)
filled_set <- select(filled_set, filled_steps, date, interval)
names(filled_set) <- names (data_set)
```

### Data set before missing values are replaced:
```{r}
data_set
```

### Data set after missing values are replaced:
```{r}
filled_set
```

###  Histogram of the total number of steps taken each day 
```{r}
#Summarize to steps per day
steps_per_day_filled <-summarize(group_by(filled_set,date),total_steps =sum(steps, na.rm = TRUE))
```


```{r}
#Histogram of number of steps per day
with(steps_per_day_filled, hist(total_steps, breaks=10))
```
### Calculate and report the mean and median total number of steps 

Mean:
```{r}
#Mean of Number of total steps per day
summarize(steps_per_day_filled,mean_steps =mean(total_steps, na.rm = TRUE))
```

Median:
```{r}
#Median Number of total steps per day
summarize(steps_per_day_filled,median_steps =median(total_steps, na.rm = TRUE))
```
### Conclusion

The mean and median values have changed as they were dragged down by the missing values being classed as a zero.
The new histogram is much smoother and closer to a nromal distribution.

## Are there differences in activity patterns between weekdays and weekends?

Firstly we flag the dates to indicate weekday or weekend and create a data set for the mean of each. These are then merged into a data set called "weekday"

```{r}
#Create extra column to indicate weekday/weekend
filled_set$weekday <-ifelse(weekdays(steps_per_day_filled$date)=="Saturday"|weekdays(steps_per_day_filled$date)=="Sunday","WEEKEND","WEEKDAY")
weekday<-arrange(summarize(group_by(filled_set[filled_set$weekday=="WEEKDAY"], interval),weekday_steps =mean(steps, na.rm = TRUE)),interval)
weekend<-arrange(summarize(group_by(filled_set[filled_set$weekday=="WEEKEND"], interval),weekend_steps =mean(steps, na.rm = TRUE)),interval)
weekday$weekend_steps <- weekend$weekend_steps
```

### This is a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)


```{r}
#Plot output comparing weekday to weekend
par(mfrow=c(2,1))

with(weekday,  plot(interval, weekday_steps, type = "n", xlab="Time Interval Number", ylab="Average Number of Steps"))
lines(weekday$interval,weekday$weekday_steps, col="blue",type ="S")
title(main = "weekday")
with(weekday,  plot(interval, weekend_steps, type = "n", xlab="Time Interval Number", ylab="Average Number of Steps"))
lines(weekday$interval,weekday$weekend_steps, col="blue",type ="S")
title(main = "weekend")
```
## Conclusion

- Cleaning up the data replace missing values gives a more accurate view of the distibution of activity.

- It is clear that although the time distribution of activity is similar between weekdays and weekends, the overall activity is higher at weekends. A plot on the same set of axes might show this more clearly.



